---
title: "Random Forest"
author: "Rahul Singh" 
date: "3/26/2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading libraries
```{r}
library(base)
library(lubridate)
library(dplyr)
library(haven)
library(data.table)
library(Matrix)
library(geosphere)
library(mltools)
library(mlr)
library(caret)
library(readr)
library(liquidSVM)
library(stringr)
library(randomForest)
library(rpart)
library(ggplot2)

```

Importing data
```{r}
rm(list=ls()) #to remove previously existing data in the R memory
master_data<- read.csv("~/Desktop/Pump_it_up_ML_2/DATASET/Training_set_values.csv")
master_data_labels<-read.csv("~/Desktop/Pump_it_up_ML_2/DATASET/Training_set_lables.csv")
test_data<-read.csv("~/Desktop/Pump_it_up_ML_2/DATASET/Test_set_values.csv")
```

Checking for NAs
```{r}
sapply(master_data, function(x) sum(is.na(x))) #we observe no missing values
```
Checking for duplicate rows
```{r}
master_data[duplicated(master_data),]  # shows which rows are repeated 
#in this case there are no duplicate rows
```

```{r}
model1_df<-master_data
test_model1_df<-test_data
```

```{r}
#Replacing all the whitespace in names of columns with underscore so that later on we do not face any issues while calling on the names
names(model1_df) <- gsub(" ", "_", names(model1_df))

#levels(model1_df$basin)

#Replacing all the whitespace in factors with underscore so that later on we do not face any issues while calling on the names
levels(model1_df$basin) <- gsub(" ", "_", levels(model1_df$basin))

levels(model1_df$region) <- gsub(" ", "_", levels(model1_df$region))

#region codes are integers; this needs to be converted to factors
model1_df$region_code<- as.factor(model1_df$region_code) #27 levels

model1_df$district_code<- as.factor(model1_df$district_code) #20 levels

#lga row is already a factor with 125 levels
#ward is also a factor with 2092 levels

#population round the well is okay as integers

#since permit column has empty values in true and false, we can split the column into three sub columns permit_true with 1 for true and 0 otherwise;
#columns permit_false with 1 for value matching false and 0 otherwise;
model1_df$permit_true<-ifelse(model1_df$permit=="True",1,0)
model1_df$permit_false<-ifelse(model1_df$permit=="False",1,0)
model1_df$permit_dontknow<-ifelse(model1_df$permit=="",1,0)

#remove the permit column
model1_df <- subset(model1_df, select = -c(permit))
```



```{r}
levels(model1_df$extraction_type) <- gsub(" ", "_", levels(model1_df$extraction_type)) #removing white space in levels

levels(model1_df$extraction_type_group) <- gsub(" ", "_", levels(model1_df$extraction_type_group))

levels(model1_df$extraction_type_class) <- gsub(" ", "_", levels(model1_df$extraction_type_class))

levels(model1_df$management) <- gsub(" ", "_", levels(model1_df$management))

#management group looks fine in terms of whitespace
levels(model1_df$payment) <- gsub(" ", "_", levels(model1_df$payment))

levels(model1_df$payment_type) <- gsub(" ", "_", levels(model1_df$payment_type))

levels(model1_df$water_quality) <- gsub(" ", "_", levels(model1_df$water_quality))

levels(model1_df$quality_group) <- gsub(" ", "_", levels(model1_df$quality_group))

#quantity has 5 factors and no whitespace

#quantity and quantity_group are the same, so we remove it
model1_df <- subset(model1_df, select = -c(quantity_group))

levels(model1_df$source) <- gsub(" ", "_", levels(model1_df$source))

#source and source_type are the same, so we remove the other column
model1_df <- subset(model1_df, select = -c(source_type))

levels(model1_df$source_class) <- gsub(" ", "_", levels(model1_df$source_class))

levels(model1_df$waterpoint_type) <- gsub(" ", "_", levels(model1_df$waterpoint_type))

#remove waterpoint_type_group as it is the same as waterpoint_type

model1_df <- subset(model1_df, select = -c(waterpoint_type_group))

model1_df$permit_true<- as.factor(model1_df$permit_true)
model1_df$permit_false<- as.factor(model1_df$permit_false)
model1_df$permit_dontknow<- as.factor(model1_df$permit_dontknow)

```


```{r}
#Replacing all the whitespace in names of columns with underscore so that later on we do not face any issues while calling on the names
names(test_model1_df) <- gsub(" ", "_", names(test_model1_df))

#levels(model1_df$basin)

#Replacing all the whitespace in factors with underscore so that later on we do not face any issues while calling on the names
levels(test_model1_df$basin) <- gsub(" ", "_", levels(test_model1_df$basin))

levels(test_model1_df$region) <- gsub(" ", "_", levels(test_model1_df$region))

#region codes are integers; this needs to be converted to factors
test_model1_df$region_code<- as.factor(test_model1_df$region_code) #27 levels

test_model1_df$district_code<- as.factor(test_model1_df$district_code) #20 levels

#lga row is already a factor with 125 levels
#ward is also a factor with 2092 levels

#population round the well is okay as integers

#since permit column has empty values in true and false, we can split the column into three sub columns permit_true with 1 for true and 0 otherwise;
#columns permit_false with 1 for value matching false and 0 otherwise;
test_model1_df$permit_true<-ifelse(test_model1_df$permit=="True",1,0)
test_model1_df$permit_false<-ifelse(test_model1_df$permit=="False",1,0)
test_model1_df$permit_dontknow<-ifelse(test_model1_df$permit=="",1,0)

#remove the permit column
test_model1_df <- subset(test_model1_df, select = -c(permit))
```



```{r}
levels(test_model1_df$extraction_type) <- gsub(" ", "_", levels(test_model1_df$extraction_type)) #removing white space in levels

levels(test_model1_df$extraction_type_group) <- gsub(" ", "_", levels(test_model1_df$extraction_type_group))

levels(test_model1_df$extraction_type_class) <- gsub(" ", "_", levels(test_model1_df$extraction_type_class))

levels(test_model1_df$management) <- gsub(" ", "_", levels(test_model1_df$management))

#management group looks fine in terms of whitespace
levels(test_model1_df$payment) <- gsub(" ", "_", levels(test_model1_df$payment))

levels(test_model1_df$payment_type) <- gsub(" ", "_", levels(test_model1_df$payment_type))

levels(test_model1_df$water_quality) <- gsub(" ", "_", levels(test_model1_df$water_quality))

levels(test_model1_df$quality_group) <- gsub(" ", "_", levels(test_model1_df$quality_group))

#quantity has 5 factors and no whitespace

#quantity and quantity_group are the same, so we remove it
test_model1_df <- subset(test_model1_df, select = -c(quantity_group))

levels(test_model1_df$source) <- gsub(" ", "_", levels(test_model1_df$source))

#source and source_type are the same, so we remove the other column
test_model1_df <- subset(test_model1_df, select = -c(source_type))

levels(test_model1_df$source_class) <- gsub(" ", "_", levels(test_model1_df$source_class))

levels(test_model1_df$waterpoint_type) <- gsub(" ", "_", levels(test_model1_df$waterpoint_type))

#remove waterpoint_type_group as it is the same as waterpoint_type

test_model1_df <- subset(test_model1_df, select = -c(waterpoint_type_group))

test_model1_df$permit_true<- as.factor(test_model1_df$permit_true)
test_model1_df$permit_false<- as.factor(test_model1_df$permit_false)
test_model1_df$permit_dontknow<- as.factor(test_model1_df$permit_dontknow)
```


```{r}
model1_df$fe_days <- as.numeric(as.Date("2014-01-01") - as.Date(model1_df$date_recorded))
model1_df$fe_mont <- month(ymd(model1_df$date_recorded))
model1_df$fe_dist <- distGeo(as.matrix(model1_df[,c('longitude','latitude')]), c(0,0))

test_model1_df$fe_days <- as.numeric(as.Date("2014-01-01") - as.Date(test_model1_df$date_recorded))
test_model1_df$fe_mont <- month(ymd(test_model1_df$date_recorded))
test_model1_df$fe_dist <- distGeo(as.matrix(test_model1_df[,c('longitude','latitude')]), c(0,0))
```

The amount variable is skewed, the log transformation can help during the quantile sketch. Addition of one is needed to avoid missing values.The square is added and a dummy to flag values above the third quantile.
```{r}
model1_df$amount <- log(model1_df$amount_tsh + 1)
model1_df$amount_squared <- model1_df$amount * model1_df$amount
model1_df$amount_q3 <- 0
model1_df$amount_q3[model1_df$amount > 3.044] <- 1
```

```{r}
test_model1_df$amount <- log(test_model1_df$amount_tsh + 1)
test_model1_df$amount_squared <- test_model1_df$amount * test_model1_df$amount
test_model1_df$amount_q3 <- 0
test_model1_df$amount_q3[test_model1_df$amount > 3.044] <- 1
```


The population is log transformed. Based on the histogram the log transformed distribution has a saddle. Data points above the saddle are flagged. 
```{r}
model1_df$population <- log(model1_df$population + 1)
model1_df$population_below <- 0
model1_df$population_below[model1_df$Population < 2] <- 1
```

```{r}
test_model1_df$population <- log(test_model1_df$population + 1)
test_model1_df$population_below <- 0
test_model1_df$population_below[test_model1_df$Population < 2] <- 1
```

```{r}
#' Finally, `lga` (local geographic area?) is interesting because there are distinct areas (e.g. *arusha*) but some of them are split into rural and urban (e.g., *arusha rural* and *arusha urban*). I transform this variable into a new feature that takes three values: rural, urban and other.

model1_df = model1_df %>% mutate(lga = ifelse( grepl(" rural", lga), "rural",ifelse( grepl(" urban", lga), "urban","other")))
test_model1_df = test_model1_df %>% mutate(lga = ifelse( grepl(" rural", lga), "rural",ifelse( grepl(" urban", lga), "urban","other")))
```

```{r}
train_data_labels<-master_data_labels["status_group"]
```


**Remember that id column is still in data**
```{r}
model1_df<-model1_df[,-c(2,3,4,6,7,8,9,10,12,17,19,20,21,22,24)]
test_model1_df<-test_model1_df[,-c(2,3,4,6,7,8,9,10,12,17,19,20,21,22,24)]
```

```{r}
columns= (colnames(test_model1_df))
train_and_test<- rbind(model1_df,test_model1_df)
end_train <- nrow(model1_df)
end <- nrow(train_and_test)
trainm <- left_join(x = model1_df, y = master_data_labels, by = "id")
testm <- test_model1_df
```

```{r}
trainm=trainm %>% mutate_if(is.character, as.factor)
testm=testm %>% mutate_if(is.character, as.factor)
```

```{r}
ntrees <- 300
```

```{r}
trainm <- trainm[,-1]
testm <- testm[,-1]
```


```{r}

model_forest <- randomForest(as.factor(status_group) 
                             ~ .,data = trainm
                             ,importance = TRUE
                             ,ntree = ntrees
                             ,nodesize = 5
)

```

```{r}
pred_forest_train<- predict(model_forest,trainm)
table(trainm$status_group,pred_forest_train)
```

```{r}
common <- intersect(names(trainm), names(testm)) 
for (p in common) { 
  if (class(trainm[[p]]) == "factor") { 
    levels(testm[[p]]) <- levels(trainm[[p]]) 
  } 
}


pred_forest_test<- predict(model_forest,testm)
submissionForest<-data.frame(test_data$id)
submissionForest$status_group<- pred_forest_test
str(submissionForest)
names(submissionForest)[1]<-"id"
names(submissionForest)[2]<-"status_group"
submissionForest$status_group <- as.factor(submissionForest$status_group)
```

```{r}
file_out <- c("Random_Forest_Model2_trial.csv")
write.csv(submissionForest, file = file_out, row.names=F,col.names=F,sep="\t", quote=FALSE)
```
